---
title: "Cyclistic_Bike_Share_Case_Study"
output:
  pdf_document: default
  word_document: default
date: "2022-07-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) 
library(lubridate) 
library(ggplot2) 
getwd() 
setwd("/Users/justi/OneDrive/Desktop/Data Science/Cyclistic Bike Share Case Study")
```

```{r}
install.packages('formatR', repos = "http://cran.rstudio.com")

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r}
#Load each quarterly dataset of the last year here
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

```{r}
#Check that the names of the datasets are consistent

colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```
```{r}
# Rename columns in 2019 to make them consistent with q1_2020, several columns were dropped in q1_2020

(q2_2019 <- rename(q2_2019, ride_id = "01 - Rental Details Rental ID",rideable_type = "01 - Rental Details Bike ID",started_at = "01 - Rental Details Local Start Time",ended_at = "01 - Rental Details Local End Time",start_station_name = "03 - Rental Start Station Name",start_station_id = "03 - Rental Start Station ID",end_station_name = "02 - Rental End Station Name",end_station_id = "02 - Rental End Station ID",member_casual = "User Type"))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))


```



```{r}
# Inspect the dataframes and look for incongruencies
str(q2_2019)
str(q3_2019)
str(q4_2019)
str(q1_2020)

```
```{r}
# Convert ride_id and rideable_type to character so that they can stack correctly

q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```

```{r}
# Stack individual quarter's data frames into one big data frame

all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

```{r}
# Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```

###Data Cleaning & Transformation

```{r}
# Inspect the new table that has been created
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)  
```

Issue #1: Within the "member_casual" column, there seem to be four different labels, Subscriber, member, customer, and casual. Based on the given information within the case study, customers should be labeled as either a member or a casual. This will be solved by renaming all subscribers as members and all customers as casual.

Addition #1: The date and time stamps for this dataset are all combined into one character string and placed in either the "started_at" or "ended_at" columns. Splitting this data up into separate columns into time, day, month, and year will allow us to make more meaningful connections in our analysis

Addition #2: Currently, there are columns that tell us the start_station id/name and the end_station id/name. This information is not quantifiable, meaning that we cannot utilize it in its raw state. To use it effectively, we will add a column called "ride_duration" that tells us how long it took to get from the start station to the end station.

```{r} 
#Issue 1
all_trips <- all_trips %>% mutate(member_casual = recode(member_casual, "Subscriber" = "member", "Customer" = "casual"))
```

```{r}
#Addition #1
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
all_trips$hour <- format(as.POSIXct(all_trips$started_at), "%H")
```


```{r}
#Addition #2
all_trips$ride_duration <- difftime(all_trips$ended_at,all_trips$started_at)
```

```{r}
#checking for sign value in ride_duration column
all_trips$sign_check <- sign(all_trips$ride_duration)

3879822 - sum(all_trips$sign_check)

#Since the above function doesn't output zero, it means that there are some zeros or negative ones within the sign check column. Because it doesn't match up with the total number of observations in our dataset, we can conclude there are negative ride durations within the column. There are also some rides where Cyclistic took the bikes out of service for maintenance. These all need to be removed and cleaned before continuing with the analysis.

```

```{r}
#Clean the bad data from the dataset & create a new dataset

all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_duration<0),]
```

###Analysis

```{r}
#mean, median, max, min of all rides
all_stats <- all_trips_v2 %>% summarize(mean = mean(ride_duration), median = median(ride_duration), max = max(ride_duration), min = min(ride_duration))
all_stats

#mean, median, max, min of members and casual customers
all_stats_members <- all_trips_v2 %>% filter(member_casual == "member") %>% summarize(mean = mean(ride_duration), median = median(ride_duration), max = max(ride_duration), min = min(ride_duration))
all_stats_members

all_stats_casual <- all_trips_v2 %>% filter(member_casual == "casual") %>% summarize(mean = mean(ride_duration), median = median(ride_duration), max = max(ride_duration), min = min(ride_duration))
all_stats_casual

#average ride time per day split by members and casual
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
avg_ride_time_per_day <- aggregate(all_trips_v2$ride_duration ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
avg_ride_time_per_day

```

```{r}
#Grouped data by member type and day of week
grouped_data_week <- all_trips_v2 %>% group_by(member_casual, day_of_week) %>% summarise(number_of_rides = n(),average_duration = mean(ride_duration), .groups = 'drop') %>% arrange(member_casual, day_of_week)
grouped_data_week

```
```{r}
#Grouped data by member type and month
grouped_data_month <- all_trips_v2 %>% group_by(member_casual, month) %>% summarise(number_of_rides = n(), .groups = 'drop') %>% arrange(member_casual, month)
grouped_data_month
```
```{r}
#Grouped data by member type and hour
grouped_data_hour <- all_trips_v2 %>% group_by(member_casual, hour) %>% summarise(number_of_rides = n(), .groups = 'drop') %>% arrange(member_casual, hour)
grouped_data_hour
```

```{r}
#Bar graph of number of bike rides per group per day (Graph #1)
bar_bikerides_per_group_per_day <- grouped_data_week %>% ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge") + labs(title = "Number of Rides per Day for each Member Type", x = "Day of Week", y = "Number of Rides")
bar_bikerides_per_group_per_day

```

```{r}
#Bar graph of average length of bike rides per group per day (Graph #2)
avg_length_bar_bikerides_per_group_per_day <- grouped_data_week %>% ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) + geom_col(position = "dodge") + labs(title = "Average Duration of Rides per Day for each Member Type", x = "Day of Week", y = "Average Duration")
avg_length_bar_bikerides_per_group_per_day
```

```{r}
#number of rides per month for each member type (Graph #3)
bar_bikerides_per_group_per_month <- grouped_data_month %>% ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge") + labs(title = "Number of Rides per Month for each Member Type", x = "Month", y = "Number of Rides")
bar_bikerides_per_group_per_month
```

```{r}
#Most frequent bike ride starting hour for each member type (Graph #4)
bar_num_bikerides_per_each_hour <- grouped_data_hour %>% ggplot(aes(x = hour, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge") + labs(title = "Number of Rides per Hour for each Member Type", x = "Hour", y = "Number of Rides")
bar_num_bikerides_per_each_hour
```

